---
title: Previous Pacific Cruise Data
weight: 9
---

I've compiled all of the previous data from our cruise track into one R package that you can download and access. This page will walk you through that data so you can access and investigate.

## Getting the data

To start, you will need to install `devtools` which are a selection of developer tools that can be used for lots of things. In our case, accessing a package stored on [GitHub](https://github.com/) rather than the standard R package repository (CRAN).

To install, you just need to run this command once (i.e. in the console, not in a script)

```{r, eval = FALSE}
install.packages("devtools")
```

Then you can download my `seapacific` package, again with a single command.

```{r, eval = FALSE}
devtools::install_github("benharden27/seapacific")
```

That will save the data package to your computer. You will be able to use the data by including that library in any script you run.

```{r}
library(seapacific)
```

## Data orientation

To use all of the data, we're going to need to make sure we have the tidyverse loaded.

```{r}
library(tidyverse)
```


There is one data file per cruise:

* S258
* S265
* S272
* S278
* S285

You can access the data by simply typing the name of the cruise. I'll do it using the `summary` function so you can see what it there.

```{r}
summary(S265)
```

Each of these fields is a subset of the total data by type. Many of them are simple data frames:

* elg: minute-by-minute surface data where available (location, time, sal, temp, etc.)
* hourly: as elg but every hour and typically available for all cruises 
* surfsamp: surface properties at stations when we took a water sample (includes N, P, chla, etc.)
* neuston: surface net tow data (zooplankton biomass, plastics, halobates, diversity, etc.)
* hydro: bottle data from a hydrocast (like surfsamp but for subsurface samples)
* ctd2: ctd data listed by cruise and station (temp, sal, etc.)

There are a couple of non-dataframe datasets which we can discuss if you need them

* adcp: velocity data from ADCP
* ctd: individual ctd data files in the `oce` format

Any of them can be accessed using the `$` symbol. For example

```{r}
S265$neuston
```

Is the neuston tow data sheet. And again, you can view this like an Excel sheet using

```{r, eval = FALSE}
View(S265$neuston)
```

Regrettably, these data are not well documented so you will have to use your best guess or ask me if you want to know what a particular field is. Hopefully they should be moderately self explanatory. Hopefully.

## Combining years

If you want to combine data from different years, here is a starting point. You probably want to give each you are using a cruise ID as a column using the mutate function.

```{r}
S258$hourly <- mutate(S258$hourly, ID = "S258")
S265$hourly <- mutate(S265$hourly, ID = "S265")
S272$hourly <- mutate(S272$hourly, ID = "S272")
S278$hourly <- mutate(S278$hourly, ID = "S278")
S285$hourly <- mutate(S285$hourly, ID = "S285")
```

Then you can combine them altogether into one big data set

```{r}
all_hourly <- bind_rows(S258$hourly, 
                        S265$hourly,
                        S272$hourly,
                        S278$hourly,
                        S285$hourly)
```

Then we can plot and use the `ID` field to group or use an an aesthetic

```{r}
ggplot(all_hourly) +
  geom_point(aes(lat, temp, color = ID))

```

