---
title: Plotting II - Changing Aesthetics
weight: 4
---

Data for this exercise can be found [here](../../data_raw/S258_surfsamp.csv).

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, include = TRUE, cache = TRUE, message = FALSE, warning = FALSE)
```

In this class, we will continue to practice plotting data using [ggplot](https://ggplot2.tidyverse.org/index.html) and extend what we know to include how we format our plots to look nice for publication. We will be plotting data from a previous SEA cruise (S258)

## Prerequisites

Before this class, you should have worked through [chapter 28](https://r4ds.had.co.nz/graphics-for-communication.html) of [R for Data Science](https://r4ds.had.co.nz/index.html) by Garrett Grolemund and Hadley Wickham. This chapter introduces you to the ideas behind how to format the parts of your plot and should be used as a reference for what we are going to do next.

Again, open a new R Script for this class (File -> New File -> R Script) to store all the code you are going to write. Don't forget to comment as it will help you go back and use this code in the future.

To complete the exercises contained below, you will need to load the group of packages in the `tidyverse`:

```{r, include = TRUE}
library(tidyverse)
```

or if you have issues with any packages contained in the tidyverse, you can load the following individual packages:

```{r, include = TRUE}
library(ggplot2)
library(readr)
```

## Reading in the data

The data we are going to be using is contained in the `surfsamp` data set that SEA records. It is a record of values for whenever we collected a surface sample of water and conducted some analysis on it. More details below. 

Just like last class, you are going to read in the data using the `read_csv()` function. You need to go get the data from the top of this page and save it somewhere sensible on your computer. As before, this is a comma-separated-file that you can check out in Excel if that is more comfortable for you.

On my computer, the read function looks like this:

```{r read_data, include = TRUE}
df <- read_csv("~/Documents/SEA/S285_OC/dcv/R/classes/data/S258_surfsamp.csv")
```

As before, you will need to replace the path I have given with the path on your computer.

## Initial look at the data

Like last time, we can get an initial look at this data in a few different ways. The following prints out the start of the data:

```{r, include = TRUE}
df
```

This, creates the standard Excel look of the data:

```{r view, include = TRUE, eval = FALSE}
View(df)
```

And we can create a summary of the data too:

```{r, include = TRUE}
summary(df)
```

As with our previous data we can see a number of rows of observations and columns of variables:

- The columns are:
    + station: Surface station identification number (SS-###)
    + dttm: Date and time (in R's special datetime format)
    + lon: Longitude (decimal degrees east of 0)
    + lat: Latitude (decimal degrees north of 0)
    + temp: Surface temperature (^o^C)
    + sal: Surface salinity (PSU)
    + fluor: Surface fluorescence (Volts - a measure of chlorophyll-a concentration)
    + chla: Extracted Chl-a Concentration (mg/L)
    + no3: Nitrate concentration (uMol/L)
    + po4: Phosphate concentration (uMol/L)
    + pH: Ocean pH
    + alk: Alkalinity of the ocean (Meq/L)
    
## Initial plots

Like last time, we can follow the ggplot recipe of combining data, mapping and geoms to create some first look plots.

```{r, include = TRUE}
ggplot(df) +
  geom_line(aes(dttm, temp))
```

This is approximately the same plot we made last time - temperature increasing as we head north along our cruise track. As before we can use different geoms to plot this data in different ways depending on how we deem it best to be represented:

```{r, include = TRUE}
ggplot(df, aes(dttm, temp)) +
  geom_point() +
  geom_smooth()
```

## Changing appearance

I'm just going to run through some of the key parts of the homework chapter now to see how we apply those concepts to our data set to take control over the appearance. There are a million different things you could do and the best way to find out new ways to do things is to think of something you'd like to do and then find a solution online to how to do it. That said, here are a few ideas of typical changes you might want to make.

### Lines and Points

You can change the colors of you plot when you define the geom. All you have do do is declare the color, size, etc. outside of the `aes()` function. That way you are saying that you want the geom to have that property universally rather than mapping a variable to that aesthetic.

```{r}
ggplot(df, aes(dttm, temp)) +
  geom_point(color = "red", size = 10, shape = 5) +
  geom_smooth(color = "black")
```

Color can also be specified in the RGB format "#RRGGBBAA". See the color brewer page for more info on that. Below is an example of some of the stuff you can change for a line.

```{r}
ggplot(df, aes(dttm, temp)) +
  geom_line(color = "#2ca25f", size = 3, linetype = "dashed", lineend = "round", linejoin = "mitre")
```


You can check out all the aesthetics you can set in the ggplot vignette below.

```{r, eval = FALSE}
vignette("ggplot2-specs")
```

### Labels

Right now, ggplot automatically labels the axis and color/size scales you draw with the names of the variables as they appear in the data frame. It is likely that you'll want to change that and you can use the `labs` function to do that - you just add it on like you would a geom. In the following example, I've also added salinity as color aesthetic and assigned the plot to a variable to reuse later:
 
```{r, include= TRUE}
g <- ggplot(df, aes(dttm, temp, color = sal)) +
       geom_point(size = 3) +
       labs(title = "Temperature verses time colored by salinity",
            x = "Time",
            y = "Temperature",
            color = "Salinity")
g
```

### Scaling

You can change the color scales on your plots using functions in ggplot starting with `scale_color_*`. Some good options include:

- Use one of the built-in color scale tools such as `scale_color_gradient()` which creates a gradient between two colors or `scale_color_gradient2()` which creates a color map between three colors

- [colorbrewer](http://colorbrewer2.org) provides a good selection of well thought-out and suitable color maps for discrete data and is implemented in ggplot using `scale_color_brewer()` for discrete data and `scale_color_distiller()` for continuous data

```{r}
g + 
  scale_color_distiller(palette = "GnBu")
```


- the `viridis` package is designed specifically for continuous data.

```{r}
library(viridis)
g +
  scale_color_viridis()
```

- the `oce` package has some typical scales used throughout oceanography

```{r}
library(oce)
g +
  scale_color_gradientn(colors = oceColorsTemperature(100))
```

You can also change the number of breaks and labels as well (NB: the following example is silly, but shows that you can replace labels with anything that you like as long as your labels vector is the same length as your breaks vector). 

```{r}
g +
  scale_color_viridis(breaks = seq(33.5,35.5,.5), labels = c("a","b","c","d","e"))
```

To that end, you can also create custom tick breaks for the x and y axes too:

```{r}
g +
  scale_y_continuous(breaks = c(16,17,20,21,25))
```


You can also curtail your color axis like you might an x and y axis

```{r, include = TRUE}
g + 
  scale_color_viridis(limits = c(35,35.5))
```

You can see in this example that all values outside of this range have become gray as they no longer exist in our limits

## Zooming

As we spoke about last week, you can either zoom within one of the `scale_` functions or within a `coord_` function. The former cuts the data out of the data set, the latter just zooms us in on a region. In most cases, you are going to want to use `coord_cartesian()` to zoom in on some part of the plot. This zooms without destroying data outside of your range which can be important for things like boxplots and smoothing lines. NB you can also use the functions `xlim()` and `ylim()` as shorthand too.

```{r, include = TRUE}
g + 
  coord_cartesian(ylim = c(20,24))
```

You can also use a different coordinate system - one we are likely to want sometime soon is `coord_quickmap()` which formats lon and lat to be well proportioned. More in a week or so!

## Themes

You can change just about anything you want about the look of your figure in the `theme()` function. Before you go to far into that though, ggplot provides some good basic themes to help you out. These are typically applied at the end of the call to ggplot:

```{r, include = TRUE}
g +
  theme_light()
```

You can also specify individual aspects of you theme using the `theme()` function. Check out the help file to see how that works and what you can change.

## Next steps / Practice

There's a lot to play around with to customize your plot. The next step is to play around with this data set and start plotting variables and playing around with how you want it all to look. I've given you some ideas below of ways to experiment with some of the features we've discussed, but at the end of the day, what you plot is up to you.

> Plot pH as a function of time along the cruise track. Add a smoothing line. Create labels and change the theme to make it look presentable. Plot temperature as color in the plot and play around with color palette options.

> Go back to the neuston data from last session, make the same plot of fluoroescence as a function of time, and color by daynight. Now our colormap is a discrete one rather than continuous - play around with your color options for this discrete scale.




